Modelling M/EEG data with Bayesian generalised additive multilevel models

### 2025-02-13 brms_meeg.qmd
pushing some preliminary simulation results

---
title: "Modelling M/EEG data with Bayesian multilevel generalised additive models"
shorttitle: "Bayesian M/EEG modelling"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
    # financial-support: "This study was supported by Grant [Grant Number] from [Funding Source]."
abstract: "Time-resolved electrophysiological measurements such as those offered by magneto- or electro-encephalography (M/EEG) provide a unique window onto neural activity underlying cognitive process and how they unfold over time. Typically, we are interested in testing whether such measures differ across conditions and/or groups. The conventional approach consists in conducting mass-univariate statistics through followed by some form of multiplicity correction (e.g., FDR, FWER) or cluster-based inference. However, these cluster-based methods have an important downside: they shift the focus of inference from the timepoint to the cluster level, thus preventing any conclusion to be made about the onset and offset of effects (e.g., differences across conditions). Here, we introduce a novel *model-based approch* for analysing one-dimensional M/EEG timeseries such as ERPs or decoding timecourses and their differences across conditions or group. This approach relies on Bayesian nonparametric multilevel modelling (multilevel generalised additive models), which outputs the posterior probabilility of the effect being above 0 (or above chance) at every timestep, while naturally taking into account the temporal dependencies and between-subject variability present in such data."
keywords: [EEG, MEG, generalised additive models, gaussian processes, mixed-effects models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# and systemfonts::system_fonts()
# pdf-engine: xelatex
# monofont: Roboto
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  # apaquarto-html: default
  apaquarto-pdf:
    latex-engine: xelatex
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
---

```{r setup, include = FALSE}
# loading packages
library(changepoint)
library(tidyverse)
library(tidybayes)
library(patchwork)
library(MetBrewer)
library(grateful)
library(scales)
library(pakret)
library(scico)
library(knitr)
library(brms)

@@

# Introduction (in progress)

Here are some useful references to be discussed [@combrisson_exceeding_2015; @ehinger_unfold_2019; @hayasaka_validating_2003; @luck_how_2017; @pedersen_hierarchical_2019; @riutort-mayol_practical_2023; @rousselet_using_2025; @frossard2021; @frossard2022; @gramfort2013; @maris2007; @pernet2015]... See also [@maris2011]... and [@rosenblatt2018] (history of cluster-based approaches and using a data split?)... Cluster failure [@eklund2016]...

In the following, we consider two approaches to modelling the non-linear timecourse of M/EEG ERPs or decoding performance: i) generalised additive models (GAMs) with thin-plate smoothing splines [@wood2003; @wood2004] and ii) Gaussian processes (GPs) with a smooth covariance kernel [@rasmussen2005] and low-rank approximation [@riutort-mayol_practical_2023].

@@

From @dimigen2021: "Recently, spline regression has been applied to ERPs (Hendrix, Baayen, & Bolger, 2017; Kryuchkova,Tucker,Wurm,&Baayen,2012;Tremblay& Baayen, 2010; Tremblay & Newman, 2015)"...

Disentangling overlapping processes [@skukies_brain_2024; @skukies_modelling_2021]... Using Bayes factors [@teichmann2022]... Weighting single trials [@pernet2022]...

See <https://www.let.rug.nl/nerbonne/teach/rema-stats-meth-seminar/presentations/Wieling-2014-GAMs.pdf> and <https://www.let.rug.nl/wieling/Statistics/GAM-EEG/lab/>, <https://www.let.rug.nl/wieling/Statistics/GAM-EEG/GAM-EEG.pdf> [@abugaber2023; @meulman2015]...

@@

@@

@@

@@

For simulating EEG data, see <https://github.com/GRousselet/onsetsim> and <https://github.com/unfoldtoolbox/UnfoldSim.jl>...

@@

We then fitted a GAM and GP regression model using the `brms` package [@brms2017; @brms2018; @nalborczyk2019]. We used the default priors in `brms`, that is, weakly informative priors. Four Markov Chain Monte-Carlo (MCMC) were ran for each model to approximate the posterior distribution, including each 5000 iterations and a warmup of 2000 iterations. Posterior convergence was assessed examining trace plots as well as the Gelman–Rubin statisti $\hat{R}$.

```{r gam, message = FALSE}
# averaging across participants
ppt_df <- raw_df %>%
    group_by(participant, condition, time) %>%
    summarise(eeg = mean(eeg) ) %>%
    # mutate(eeg = as.numeric(scale(x = eeg, scale = FALSE) ) ) %>%
    ungroup()

@@

# fitting the GAM
gam <- brm(
    # cubic regression splines with k-1 basis functions
    eeg ~ condition + s(time, bs = "cr", k = 10, by = condition),
    data = ppt_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 4,
    cores = 4,
    file = "models/gam.rds"
    )
```

Now we fit the GP regression model...

```{r gp, message = FALSE}
# computing eeg_diff and averaging across participants
# group_df <- raw_df %>%
#     group_by(participant, time) %>%
#     pivot_wider(names_from = condition, values_from = eeg, values_fn = mean) %>%
#     mutate(eeg_diff = cond2 - cond1) %>%
#     summarise(eeg_diff = mean(eeg_diff) ) %>%
#     ungroup()

@@

And we plot the posterior predictions...

```{r gam-preds, echo = FALSE}
# defining a function to plot posterior predictions
plot_post_preds <- function (model, data = NULL) {

@@

```

```{r gam-gp-preds, fig.width = 8, fig.asp = 0.5, fig.cap = "Posterior predictions of the GAM (left) and GP (right) models."}
# plotting the posterior predictions
plot_post_preds(model = gam) + plot_post_preds(model = gp_model)

```

@@

We can retrieve the posterior probability of the slope for `condition` at each timestep (@fig-plot-post-slope).

```{r fig-plot-post-slope, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Slope for the difference between conditions according to the GAM (left) or the GP (right)."}
# defining a function to plot posterior slope
plot_post_slope <- function (model, data = NULL, eps = 0.1) {

@@

    # plotting it
    prob_y_above_0 %>%
        ggplot(aes(x = time, y = m) ) +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_vline(xintercept = true_onset, linetype = 2) +
        geom_line() +
        geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
        scale_colour_met_d(name = "Johnson") +
        scale_fill_met_d(name = "Johnson") +
        labs(x = "Time (s)", y = expression(Pr(beta~"|"~data, model) ) )

@@

# using the function with model
plot_post_slope(model = gam, eps = 0.1) + plot_post_slope(model = gp_model, eps = 0.1)
```

We can also compute the posterior probability of the slope for `condition` being above $0+\epsilon$ (@fig-post-prob-test), with $\epsilon$ defined as 10% of the standard deviation of the raw EEG signal.

```{r fig-post-prob-test, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Posterior probability of the ERP difference (slope) being above 0 according to the GAM (left) and the GP (right)."}
# defining a function to plot posterior prob of slope above 0
plot_post_test <- function (model, data = NULL, eps = 0.1, n_approx = 1e3) {

@@

    # defining a sequence of time values to make predictions
    time_seq <- crossing(
        time = seq(min(data$time), max(data$time), length.out = n_approx),
        condition = c("cond1", "cond2")
        ) %>%
        arrange(condition)

    # computing the difference between cond2 and cond1
    posterior_samples <- posterior_samples[, (nrow(time_seq)/2+1):nrow(time_seq)] -
        posterior_samples[, 1:(nrow(time_seq)/2)]

    # computing the probability that y is above 0
    prob_y_above_0 <- data.frame(
        time = time_seq$time,
        m = colMeans(posterior_samples > (0 + eps) )
        # lower = apply(
        #     X = posterior_samples > (0 + eps),
        #     MARGIN = 2, FUN = quantile, probs = 0.025
        #     ),
        # upper = apply(
        #     X = posterior_samples > (0 + eps),
        #     MARGIN = 2, FUN = quantile, probs = 0.975
        #     )
        # )
        )

@@

# using the function with the two models
plot_post_test(model = gam, eps = 0.1) +
    plot_post_test(model = gp_model, eps = 0.1)
```

@@

```{r fig-post-prob-ratio, echo = FALSE, fig.width = 8, fig.asp = 0.75, fig.cap = "Ratio of posterior probability according to the GAM. Timesteps above threshold (10) are highlighted in green."}
# defining some parameters
n_approx <- 1e3
eps <- 0.1 * sd(ppt_df$eeg)
# eps <- 0.1
threshold <- 10
true_peak <- 0.250

# defining a sequence of x values to make predictions
time_seq <- crossing(
    time = seq(min(ppt_df$time), max(ppt_df$time), length.out = n_approx),
    condition = c("cond1", "cond2")
    ) %>%
    arrange(condition)

# retrieving posterior samples
posterior_samples <- posterior_epred(object = gam, newdata = time_seq)

# computing the difference between cond2 and cond1
posterior_samples <- posterior_samples[, (ncol(posterior_samples)/2+1):ncol(posterior_samples)] -
    posterior_samples[, 1:(ncol(posterior_samples)/2)]

# computing the probability that y is above 0
prob_y_above_0 <- data.frame(
    time = seq(min(ppt_df$time), max(ppt_df$time), length.out = n_approx)
    ) %>%
    mutate(post_prob = colMeans(posterior_samples) ) %>%
    mutate(m = colMeans(posterior_samples > eps) ) %>%
    mutate(prob_ratio = m / (1 - m) ) %>%
    mutate(prob_ratio = ifelse(is.infinite(prob_ratio), 1e4, prob_ratio) ) %>%
    mutate(prob_ratio = ifelse(prob_ratio == 0, 0.0001, prob_ratio) )

# printing the identified clusters
# the signal was a truncated Gaussian defining an objective onset at 160 ms,
# a maximum at 250 ms, and an offset at 342 ms.
exceeding_times_gam <- prob_y_above_0 %>%
    dplyr::filter(prob_ratio > threshold) %>%
    summarise(cluster_onset = min(time), cluster_offset = max(time) ) %>%
    mutate(
        cluster_peak = prob_y_above_0 %>%
            dplyr::filter(post_prob == max(post_prob) ) %>%
            pull(time)
        ) %>%
    mutate(true_onset = true_onset, true_peak = true_peak, true_offset = true_offset) %>%
    mutate(model = "GAM")

# creates grid for background gradient
# Create a grid covering the range of time and log-transformed y-values
bg_data <- expand.grid(
    time = seq(min(prob_y_above_0$time), max(prob_y_above_0$time), length.out = 100),
    # prob_ratio = 10^seq(log10(0.0001), log10(1e4), length.out = 100)
    prob_ratio = 10^seq(log10(min(prob_y_above_0$prob_ratio) ), log10(max(prob_y_above_0$prob_ratio) ), length.out = 100)
    )

@@

# plotting it
gam_ratio_plot <- prob_y_above_0 %>%
# prob_y_above_0 %>%
    mutate(above_thres = ifelse(prob_ratio > threshold, 1, NA) ) %>%
    ggplot(aes(x = time, y = prob_ratio) ) +
    geom_tile(
        data = bg_data,
        aes(x = time, y = prob_ratio, fill = fill_value),
        alpha = 0.4,
        inherit.aes = FALSE,
        show.legend = FALSE
        ) +
    scale_fill_scico(palette = "vik", midpoint = log10(1) ) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_hline(yintercept = 3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 1/3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 1/10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 100, linetype = "dashed", color = "orangered") +
    geom_hline(yintercept = 1/100, linetype = "dashed", color = "orangered") +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_line(linewidth = 0.8) +
    geom_point(
        data = . %>% dplyr::filter(!is.na(above_thres) ),
        aes(y = threshold),
        colour = "darkgreen",
        shape = 16,
        size = 3,
        na.rm = TRUE
        ) +
    # scale_y_log10(labels = label_log(digits = 2), limits = c(0.0001, 1e4) ) +
    scale_y_log10(
        labels = label_log(digits = 2),
        breaks = c(1/1e4, 1/1e3, 1/1e2, 1/10, 1, 10, 1e2, 1e3, 1e4),
        limits = c(0.0001, 1e4)
        ) +
    coord_cartesian(expand = FALSE) +
    labs(
        x = "Time (s)",
        y = expression(Pr(beta>0~"|"~data) / (1 - Pr(beta>0~"|"~data) ) )
        )

@@

# plotting results for the two models
# gam_ratio_plot + gp_ratio_plot
gam_ratio_plot

@@

# fitting the GAM
meta_gam <- brm(
    # using by-participant SD of ERPs across trials
    eeg_mean | se(eeg_sd) ~
        condition + s(time, bs = "cr", k = 10, by = condition) +
        (1 | participant),
    data = summary_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 4,
    cores = 4,
    file = "models/meta_gam.rds"
    )
```

```{r meta-gp, message = FALSE}
# fitting the GP
meta_gp <- brm(
    # using by-participant SD of ERPs across trials
    eeg_mean | se(eeg_sd) ~
        condition + gp(time, k = 20, by = condition) +
        (1 | participant),
    data = summary_df,
    family = gaussian(),
    control = list(adapt_delta = 0.99),
    iter = 5000,
    chains = 4,
    cores = 4,
    file = "models/meta_gp.rds"
    )
```

```{r meta-gp-preds, fig.width = 9, fig.asp = 0.5, fig.cap = "Posterior predictions for the hierarchical GAM and GP."}
# plotting the posterior predictions
meta_gam_preds <- plot(
    conditional_effects(x = meta_gam, effect = "time:condition"),
    points = FALSE, theme = theme_light(), plot = FALSE
    )[[1]] +
    scale_colour_met_d(name = "Johnson") +
    scale_fill_met_d(name = "Johnson") +
    labs(x = "Time (s)", y = "EEG signal (a.u.)")
meta_gp_preds <- plot(
    conditional_effects(x = meta_gp, effect = "time:condition"),
    points = FALSE, theme = theme_light(), plot = FALSE
    )[[1]] +
    scale_colour_met_d(name = "Johnson") +
    scale_fill_met_d(name = "Johnson") +
    labs(x = "Time (s)", y = "EEG signal (a.u.)")

# combining the two plots
meta_gam_preds + meta_gp_preds
```

@@

## Application to actual M/EEG data

Assessing the reliability of the proposed approach (in comparison to other methods) using some sort of split-half reliability [@rosenblatt2018]?

@@

We considered five methods for multiplicity corrections: two FDR methods; two cluster-based methods; and the maximum statistics. The two FDR methods were BH95 (Benjamini & Hochberg, 1995; 10.1111/j.2517-6161.1995.tb02031.x) and BY01 (Benjamini & Yekutieli, 2001; 10.1214/aos/1013699998), which were applied to the permutation p-values using the `p.adjust` function in `R`. The first cluster-based inference was implemented using a cluster-sum statistic of squared t-values...

We also compared the performance of the GAMM (in correctly identifying/estimating the onset and offset of ERP difference across conditions) to the performance of the binary segmentation method as implemented in `r pkrt("changepoint")`, threshold-free cluster enhancement [TFCE, @smith2009], as implemented in `MNE-Python` [@gramfort2013], and using the raw t-, p-, or s-values timecourse. S-values [@greenland2019]...

<!--

Comparing performance of the GAM and GP models with:

-   uncorrected univariate tests
-   univariate tests + Bonferroni/Holm
-   univariate tests + BH95 (Benjamini-Hochberg, 1995) aka FDR
-   univariate tests + BY01 (Benjamini-Yekutieli, 2001)
-   cluster-based permutation test (cluster sum)
-   cluster-based permutation test (cluster depth)
-   threshold-free cluster enhancement
-   change point

We used `r pkrt("changepoint")`...

-->

@@

```{r test-1d, echo = FALSE, fig.width = 9, fig.asp = 0.5, fig.cap = "Timecourse of squared t-values and s-values (continuous measure of evidence given by -log2(p-value)) with true onset and onset identified using the raw (uncorrected) p-values or the corrected p-values (BH, BY, Holm)."}
# https://github.com/GRousselet/onsetsim/blob/main/docs/weakstrongfwer_demo.md
# BH95: p.adjust(pvals, method = "BH")
# BY01: p.adjust(perm.pvals, method = "BY")

# defining an arbitrary alpha threshold
aath <- 0.05

>>>>

---
title: "Modelling M/EEG data with Bayesian generalised additive multilevel models"
shorttitle: "Bayesian M/EEG modelling"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
    # financial-support: "This study was supported by Grant [Grant Number] from [Funding Source]."
abstract: "Time-resolved electrophysiological measurements such as those offered by magneto- or electro-encephalography (M/EEG) provide a unique window onto neural activity underlying cognitive process and how they unfold over time. Typically, we are interested in testing whether such measures differ across conditions and/or groups. The conventional approach consists in conducting mass-univariate statistics through followed by some form of multiplicity correction (e.g., FDR, FWER) or cluster-based inference. However, these cluster-based methods have an important downside: they shift the focus of inference from the timepoint to the cluster level, thus preventing any conclusion to be made about the onset and offset of effects (e.g., differences across conditions). Here, we introduce a novel *model-based approch* for analysing one-dimensional M/EEG timeseries such as ERPs or decoding timecourses and their differences across conditions or group. This approach relies on Bayesian generalised additive multilevel models, which output the posterior probabilility of the effect being above 0 (or above chance) at every timestep, while naturally taking into account the temporal dependencies and between-subject variability present in such data."
keywords: [EEG, MEG, generalised additive models, mixed-effects models, multilevel models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# and systemfonts::system_fonts()
# pdf-engine: xelatex
# monofont: Roboto
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  # apaquarto-html: default
  apaquarto-pdf:
    latex-engine: xelatex
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
---

```{r setup, include = FALSE}
# loading packages
library(changepoint)
library(tidyverse)
library(tidybayes)
library(patchwork)
library(MetBrewer)
library(grateful)
library(ggrepel)
library(scales)
library(pakret)
library(scico)
library(knitr)
library(brms)

@@

# Introduction

Here are some useful references to be discussed [@combrisson_exceeding_2015; @ehinger_unfold_2019; @hayasaka_validating_2003; @luck_how_2017; @pedersen_hierarchical_2019; @frossard2021; @frossard2022; @gramfort2013; @maris2007; @pernet2015]... See also [@maris2011]... and [@rosenblatt2018] (history of cluster-based approaches and using a data split?)... Cluster failure [@eklund2016]...

In the following, we consider two approaches to modelling the non-linear timecourse of M/EEG ERPs (or decoding performance): i) generalised additive models (GAMs) with smoothing splines [@wood2003; @wood2004]...

and ii) Gaussian processes (GPs) with a smooth covariance kernel [@rasmussen2005] and low-rank approximation [@riutort-mayol_practical_2023].

@@

@@

From @dimigen2021: "Recently, spline regression has been applied to ERPs (Hendrix, Baayen, & Bolger, 2017; Kryuchkova,Tucker, Wurm, & Baayen,2012; Tremblay & Baayen, 2010; Tremblay & Newman, 2015)"... GAMMs for EEG data [@abugaber2023; @meulman2015]...

Disentangling overlapping processes [@skukies_brain_2024; @skukies_modelling_2021]... Weighting single trials [@pernet2022]... The LIMO toolbox (https://onlinelibrary.wiley.com/doi/10.1155/2011/831409)... Using Bayes factors [@teichmann2022]...

@@

See for instance these tutorials [@sóskuthy2017; @winter2016] or application to phonetic data [@wieling2018] or this introduction [@baayen2020] or these reference books [@hastie2017; @wood2017]...

@@

Focusing on identifying onset and offset of effects (as assessed by ERP differences or decoding performance)... Assessing the performance of a model-based approach (i.e., Bayesian GAMMs) to conventional methods (multiplicity corrections or cluster-based permutation)...

@@

@@

We then fitted a Bayesian GAM using the `brms` package [@brms2017; @brms2018; @nalborczyk2019]. We used the default priors in `brms`, that is, weakly informative priors. We ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior distribution, including each 5000 iterations and a warmup of 2000 iterations, yielding a total of $8 \times (5000-2000) = 24000$ posterior samples to use for inference. Posterior convergence was assessed examining trace plots as well as the Gelman–Rubin statistic $\hat{R}$.

```{r gam, message = FALSE}
# averaging across participants
ppt_df <- raw_df %>%
    group_by(participant, condition, time) %>%
    summarise(eeg = mean(eeg) ) %>%
    ungroup()

@@

# fitting the GAM
gam <- brm(
    # cubic regression splines with k-1 basis functions
    eeg ~ condition + s(time, bs = "cr", k = 10, by = condition),
    data = ppt_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 8,
    cores = 8,
    file = "models/gam.rds"
    )
```

```{r gp, echo = FALSE, eval = FALSE, message = FALSE}
# Now we fit the GP regression model...
# computing eeg_diff and averaging across participants
# group_df <- raw_df %>%
#     group_by(participant, time) %>%
#     pivot_wider(names_from = condition, values_from = eeg, values_fn = mean) %>%
#     mutate(eeg_diff = cond2 - cond1) %>%
#     summarise(eeg_diff = mean(eeg_diff) ) %>%
#     ungroup()

@@

```{r model-preds, echo = FALSE}
# defining a function to plot posterior predictions
plot_post_preds <- function (model, data = NULL) {

@@

}
```

```{r gam-preds, echo = FALSE, eval = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Posterior predictions of the GAM (left) and GP (right) models."}
# plotting the posterior predictions
plot_post_preds(model = gam) + plot_post_preds(model = gp_model)
```

@@

We then plot the posterior predictions together with the posterior estimate of the slope for `condition` at each timestep (@fig-plot-post-slope).

```{r fig-plot-post-slope, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Posterior estimate of the ERP in each condition (left) or directly for the difference of ERPs (right) according to the GAM."}
# defining a function to plot posterior slope
plot_post_slope <- function (model, data = NULL) {

@@

    # plotting it
    prob_y_above_0 %>%
        ggplot(aes(x = time, y = m) ) +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_vline(xintercept = true_onset, linetype = 2) +
        geom_line() +
        geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
        scale_colour_met_d(name = "Johnson") +
        scale_fill_met_d(name = "Johnson") +
        labs(
            x = "Time (s)",
            y = expression(p(beta~"|"~data, model) )
            )

@@

# plotting it
# plot_post_slope(model = gam, eps = 0.1) + plot_post_slope(model = gp_model, eps = 0.1)
plot_post_preds(model = gam) + plot_post_slope(model = gam)
```

We then compute the posterior probability of the slope for `condition` being above $0+\epsilon$ (@fig-post-prob-test), with $\epsilon := 0.05$, which can be interpreted as the smallest effect size of interest.

```{r fig-post-prob-test, echo = FALSE, out.width = "75%", fig.asp = 0.5, fig.cap = "Posterior probability of the ERP difference (slope) being above 0 according to the GAM."}
# defining a function to plot posterior prob of slope above 0
plot_post_test <- function (model, data = NULL, eps = 0.1, n_approx = 1e3, n_draws = 1e3) {

@@

    # defining a sequence of time values to make predictions
    time_seq <- crossing(
        time = seq(min(data$time), max(data$time), length.out = n_approx),
        # time = unique(data$time),
        condition = c("cond1", "cond2")
        ) %>%
        arrange(condition)

    # retrieving posterior samples
    # posterior_samples <- posterior_epred(object = model, newdata = time_seq)

    # computing the difference between cond2 and cond1
    # posterior_samples <- posterior_samples[, (nrow(time_seq)/2+1):nrow(time_seq)] -
    #     posterior_samples[, 1:(nrow(time_seq)/2)]

    # computing the probability that y is above 0
    # prob_y_above_0 <- data.frame(
    #     time = time_seq$time,
    #     m = colMeans(posterior_samples > (0 + eps) )
    #     )

    # computing probability metrics with the current eps value
    prob_y_above_0 <- time_seq %>%
        add_epred_draws(object = model, ndraws = n_draws) %>%
        data.frame() %>%
        dplyr::select(time, condition, .epred, .draw) %>%
        pivot_wider(names_from = condition, values_from = .epred) %>%
        mutate(epred_diff = cond2 - cond1) %>%
        group_by(time) %>%
        summarise(m = mean(epred_diff > 0 + eps) ) %>%
        ungroup()

@@

# using the function with the two models
plot_post_test(model = gam, eps = 0.1)
```

@@

```{r fig-post-prob-ratio, echo = FALSE, fig.width = 8, fig.asp = 0.75, fig.cap = "Ratio of posterior probability according to the GAM. Timesteps above threshold (10) are highlighted in green."}
# defining some parameters
n_approx <- 1e3
eps <- 0.05
threshold <- 10

# defining a sequence of x values to make predictions
time_seq <- crossing(
    time = seq(min(ppt_df$time), max(ppt_df$time), length.out = n_approx),
    # time = unique(ppt_df$time),
    condition = c("cond1", "cond2")
    ) %>%
    arrange(condition)

# computing the posterior probability that y is above 0
n_posterior_samples <- 24000
prob_y_above_0 <- time_seq %>%
    add_epred_draws(object = gam) %>%
    data.frame() %>%
    dplyr::select(time, condition, .epred, .draw) %>%
    pivot_wider(names_from = condition, values_from = .epred) %>%
    mutate(epred_diff = cond2 - cond1) %>%
    group_by(time) %>%
    summarise(m = mean(epred_diff > 0 + eps) ) %>%
    ungroup() %>%
    mutate(prob_ratio = m / (1 - m) ) %>%
    # replacing infinite values by maximum possible values (number of posterior samples)
    mutate(prob_ratio = ifelse(is.infinite(prob_ratio), n_posterior_samples, prob_ratio) ) %>%
    mutate(prob_ratio = ifelse(prob_ratio == 0, 1/n_posterior_samples, prob_ratio) )

# printing the identified clusters
# the signal was a truncated Gaussian defining an objective onset at 160 ms,
# a maximum at 250 ms, and an offset at 342 ms.
# exceeding_times_gam <- prob_y_above_0 %>%
#     dplyr::filter(prob_ratio > threshold) %>%
#     summarise(cluster_onset = min(time), cluster_offset = max(time) ) %>%
#     # mutate(
#     #     cluster_peak = prob_y_above_0 %>%
#     #         dplyr::filter(post_prob == max(post_prob) ) %>%
#     #         pull(time)
#     #     ) %>%
#     # mutate(true_onset = true_onset, true_peak = true_peak, true_offset = true_offset) %>%
#     mutate(true_onset = true_onset, true_offset = true_offset) %>%
#     mutate(model = "GAM")

# creating a grid covering the range of time and log-transformed y-values
bg_data <- expand.grid(
    time = seq(min(prob_y_above_0$time), max(prob_y_above_0$time), length.out = 100),
    # prob_ratio = 10^seq(log10(0.0001), log10(1e4), length.out = 100)
    prob_ratio = 10^seq(log10(min(prob_y_above_0$prob_ratio) ), log10(max(prob_y_above_0$prob_ratio) ), length.out = 100)
    )

@@

# plotting it
gam_ratio_plot <- prob_y_above_0 %>%
    mutate(above_thres = ifelse(prob_ratio > threshold, 1, NA) ) %>%
    ggplot(aes(x = time, y = prob_ratio) ) +
    geom_raster(
        data = bg_data,
        aes(x = time, y = prob_ratio, fill = fill_value),
        alpha = 0.5,
        interpolate = FALSE,
        inherit.aes = FALSE,
        show.legend = FALSE
        ) +
    scale_fill_scico(palette = "vik", midpoint = log10(1) ) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_hline(yintercept = 3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 1/3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 1/10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 100, linetype = "dashed", color = "orangered") +
    geom_hline(yintercept = 1/100, linetype = "dashed", color = "orangered") +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_line(linewidth = 0.8) +
    geom_point(
        data = . %>% dplyr::filter(!is.na(above_thres) ),
        aes(y = threshold),
        colour = "darkgreen",
        shape = 16,
        size = 3,
        na.rm = TRUE
        ) +
    scale_y_log10(
        labels = label_log(digits = 2),
        breaks = c(1/1e4, 1/1e3, 1/1e2, 1/10, 1, 10, 1e2, 1e3, 1e4)
        # limits = c(0.0001, 1e4)
        ) +
    coord_cartesian(expand = FALSE) +
    labs(
        x = "Time (s)",
        y = expression(Pr(beta>0~"|"~data) / (1 - Pr(beta>0~"|"~data) ) )
        )

@@

# plotting results for the two models
# gam_ratio_plot + gp_ratio_plot

# or plotting only the GAM
gam_ratio_plot

@@

# fitting the GAM
meta_gam <- brm(
    # using by-participant SD of ERPs across trials
    eeg_mean | se(eeg_sd) ~
        condition + s(time, bs = "cr", k = 10, by = condition) +
        (1 | participant),
    data = summary_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 8,
    cores = 8,
    file = "models/meta_gam.rds"
    )
```

```{r meta-gp, echo = FALSE, eval = FALSE, message = FALSE}
# fitting the GP
meta_gp <- brm(
    # using by-participant SD of ERPs across trials
    eeg_mean | se(eeg_sd) ~
        condition + gp(time, k = 20, by = condition) +
        (1 | participant),
    data = summary_df,
    family = gaussian(),
    control = list(adapt_delta = 0.99),
    iter = 5000,
    chains = 4,
    cores = 4,
    file = "models/meta_gp.rds"
    )
```

```{r meta-preds, echo = FALSE, fig.width = 9, fig.asp = 0.5, fig.cap = "Posterior predictions for the hierarchical GAM."}
# plotting the posterior predictions
meta_gam_preds <- plot(
    conditional_effects(x = meta_gam, effect = "time:condition"),
    points = FALSE, theme = theme_light(), plot = FALSE
    )[[1]] +
    scale_colour_met_d(name = "Johnson") +
    scale_fill_met_d(name = "Johnson") +
    labs(x = "Time (s)", y = "EEG signal (a.u.)")
# meta_gp_preds <- plot(
#     conditional_effects(x = meta_gp, effect = "time:condition"),
#     points = FALSE, theme = theme_light(), plot = FALSE
#     )[[1]] +
#     scale_colour_met_d(name = "Johnson") +
#     scale_fill_met_d(name = "Johnson") +
#     labs(x = "Time (s)", y = "EEG signal (a.u.)")

# combining the two plots
# meta_gam_preds + meta_gp_preds
```

@@

@@

We compared the ability of the GAMM to correctly estimate the onset and offset of the ERP difference to widely used methods. First, we conducted mass-univariate t-tests (thus treating each timestep independently) and identified the onset and offset of the ERP difference as the first and last values crossing an arbitrary significance threshold ($\alpha = 0.01$). We then followed the same approach but after applying different forms of multiplicity correction the the p-values. We compared two methods that control the false discovery rate (FDR) [i.e., BH95, @benjamini1995; and BY01, @benjamini2001], one method that controls the familywise error rate (FWER) (i.e., Holm–Bonferroni method, https://www.jstor.org/stable/4615733), and two cluster-based methods [permutation with a single threshold and TFCE, @smith2009]. The BH95, BY01, and Holm corrections were applied to the p-values using the `p.adjust()` function in `R`. The cluster-based inference was implemented using a cluster-sum statistic of squared t-values, as implemented in `MNE-Python` [@gramfort2013], called via `r pkrt("reticulate")`. We also compared these estimates to the onset and offset as estimated using the binary segmentation algorithm, as implemented in `r pkrt("changepoint")`, and applied directly to the squared t-values [as in @rousselet_using_2025]. For visualisation and interpretability purposes, we converted p-values to s-values, which can be interpreted as bits of surprising information, assuming a null effect [@greenland2019] (@fig-corrections).

@@

```{r fig-corrections, echo = FALSE, fig.width = 9, fig.asp = 0.5, fig.cap = "Timecourse of squared t-values and s-values, with true onset (black dashed line) and onsets identified using the raw (uncorrected) p-values or the corrected p-values (BH, BY, Holm)."}
# defining an arbitrary alpha threshold
aath <- 0.01
